---
title: "Composition Estimates with MCMC (ad-intact or no ad-clip information)"
author: "Thomas Delomas"
date: "9-16-2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Composition_Estimates_MCMC_ad_intact}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(devMCpbt)
```

## Introduction
This document will walk through how to estimate composition of a mixture of all ad-intact fish by PBT and GSI groups using a Bayesian approach. **if you do not separate wild origin fish by GSI assignments, you can create a dummy variable that gives all fish  (including PBT-assigned fish) the same GSI assignment**
This may be encountered in multiple scenarios, such as:

* The entire population you are sampling is ad-intact
* You estimate population sizes separately for ad-intact and ad-clipped fish
* You do not differentiate between ad-intact and ad-clipped (so you can analyze them as if they are all ad-intact)
* The species does not have an adipose fin (so you can analyze them as if they are all ad-intact)

This document will work through a scenario where the samples are grouped into three strata.

## First, get some data
First, we will generate some data using the built in data-simulation function.

```{r}

# define the unknown "true" proportions at which fish in each
# PBT group assign to the different GSI groups
pbtGSImat <- matrix(c(.1, .8, .1,.8, .1, .1,.1, .1, .8), nrow = 3, ncol = 3, byrow = TRUE)

# now generate three strata from identical parameters
multStratData <- data.frame()
for(i in 1:3){
	tempData <- generatePBTGSIdata(sampRate = .2, censusSize = 3000, relSizePBTgroups = c(1,2,3), tagRates = c(.8, .85,.9), physTagRates = 0, true_clipped = 0, true_noclip_H = .4, true_wild = .6, relSizeGSIgroups = c(1,2,1), PBT_GSI_calls = pbtGSImat)
	trapData <- tempData[[1]]
	trapData$Strata <- i
	multStratData <- rbind(multStratData, trapData)
}
tags <- tempData[[2]] # get the tag rates for the PBT groups
multStratData$GSI <- paste0("GSIgroup_", multStratData$GSI)

head(multStratData)
```

Now we have a dataframe named `multStratData` with a column describing whether a fish is ad-clipped or ad-intact 
(values of AD and AI, respectively), a column denoting if the fish has a physical tag that 
identifies it as hatchery origin (ignore this column for this vignette, the data we generated
has no physical tags), a column with the 
PBT assignment (Unassigned means it was attempted to be assigned, but did not have a match),
a column with GSI assignments, and a column giving the strata that sample belongs to.

```{r}
head(tags)
```

We also have a dataframe named `tags` with the names of the PBT groups (matching the labels in the 
GenParentHatchery column) and the tag rates (successful genotyping rates) for those
groups. 


## Second, prep the data for the estimation function
Now we need to prepare the data in the way the estimation function can use it. We will
do this using a function in the package that takes the two dataframes we created above
as input.

```{r}
# note that we are using AI=TRUE because we are looking at the composition
#  of ad-intact fish
mainInput <- prepStrata(multStratData, tags, GSIcol = "GSI", PBTcol = "GenParentHatchery", strataCol = "Strata", adFinCol = "AdClip", AI = TRUE)
```

This has created an input list that has one item for each strata. Each item (strata) has
the same components, just with different data. This is an important point, because the
priors can be set to be different for each strata, if desired (and if you want all the
priors changed, you must do it for all strata). Here is the structure of the input for the first strata:

```{r}
str(mainInput[[1]])
```

*For details on the structure of the input, please see the vignette titled "Structure of input and output of MCMC wrapper functions"*


##Third, we generate the estimates

We now generate the estimates of composition for each strata. These are estimates of the
proportion of fish in each strata that belong to different groups, in our case PBT and
GSI groups.



```{r}
# specifying a seed to make the anlaysis reproducible
# running  iterations with  burn-in and no thinning
# this will yield  recorded MCMC samples
propEstimates <- estimStrataMCpbt(mainInput, iter = 3000, burnIn = 500, thin = 1, seed = 7)
```

The results are structured as a list for each strata:

```{r}
str(propEstimates)
```

We can check for convergence through multiple means, such as plotting the estimates for our parameters of interest. 
Here, we will plot a few of the stock compositions in strata 1 to demonstrate:

```{r, fig.height = 3, fig.width = 7}
plot(propEstimates[[1]]$piTot[,1])
plot(propEstimates[[1]]$piTot[,5])
```

*For details on the structure of the results, please see the vignette titled "Structure of input and output of MCMC wrapper functions"*


## Fourth, we multiply by the estimated population size

In some situations, you will have a population size estimate with a particular distribution
you can sample from, while in other cases you will just have a known (or assumed) population
size. In this case, we will assume a population size of 3000 in strata 1 and 2, but a size of
1500 in strata 3. We turn this into an input the function can use as such:

```{r}
nPost <- nrow(propEstimates[[1]]$piTot)
popSizeEstimates <- list() # a list with one entry per strata
popSizeEstimates <- list(rep(3000, nPost),
					  rep(3000, nPost),
					  rep(1500, nPost))
# note that if our population size was an estimate from a given distribution,
#   we would take nPost samples from this distribution for each strata, and
#   organize them as a list, just as we are doing here
```

Now that we have the population size input, we can use a function in the package
to multiply the population sizes by the estimated proportions. If we use the 
`writeSummary = TRUE` option, a group of summary files with CI's will be
written to the working directory.

```{r}
countEstimates <- multByEscapement(mainInput, mainRes = propEstimates, popSizes = popSizeEstimates, writeSummary = TRUE)
```



Now we have two files in the working directory, one summarizing the estimated number of 
hatchery and wild fish, and the other summarizing the estimated numbers of each PBT and
wild-GSI group. The CI's in these files are symmetric CI's calculated using R's
`quantile` function. We can read them into R and take a look at them if we want:

```{r}
rearType <- read.table("RearTypeSummary.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
stockComp <- read.table("StockCompSummary.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
head(rearType)
head(stockComp)
file.remove(c("RearTypeSummary.txt", "StockCompSummary.txt")) #clean up unneeded output files
```

The variable `countEstimates` has the raw estimates calculated by multiplying the 
samples from the posterior as appropriate, and can be used to calculate other quantities
of interest or to calculate CIs using other methods. 

```{r}
str(countEstimates)
```

*For details on the structure of the results, please see the vignette titled "Structure of input and output of MCMC wrapper functions"*
